Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes

### 2023-11-07 Journal-manuscript/Journal-manuscript.tex
Edits to the introduction

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multimodal spatial availability: a singly-constrained
measure of competitive accessibility considering multiple
modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anastasia Soukhov\textsuperscript{\getAff{McMaster}}\textsuperscript{*},
Javier Tarriño-Ortiz\textsuperscript{\getAff{Madrid}},
Julio A. Soria-Lara\textsuperscript{\getAff{Madrid}},
Antonio Páez\textsuperscript{\getAff{McMaster}}\\
\bigskip
\textbf{\getAff{McMaster}}Department of Earth, Environment and Society,
McMaster University, Hamilton, Canada\\
\textbf{\getAff{Madrid}}Centro de Investigación del Transporte
(TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain\\
\bigskip
* Corresponding author: soukhoa@mcmaster.ca (AS)\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
An increasing number of studies within the domain of transportation
planning are concerned with the inequities in accessibility to
opportunities. A dimension of these inequities arises from differences
in access by mode type (e.g., commuting using a car as opposed to
transit). However, methods implemented in current accessibility
literature are lacking within the context of multimodal analysis. This
paper presents an extension of spatial availability, a
singly-constrained competitive accessibility measure, for the context of
multimodal accessibility analysis. We first illustrate the features of
spatial availability that lend itself to multimodal analysis. We then
demonstrate its use on the case study of Low Emission Zones in Madrid
(Spain) and highlight how this policy intervention changes the
accessibility of populations using different modes. In summary, spatial
availability can be used to create and interpret multimodal policy
intervention scenarios unlike previous methods: this creation and
interpretation can help regions envision a more sustainable and
equitable access-to-opportunity landscape.

@@

Implementing urban policies that re-shape cities through accessibility
gains (i.e., the \emph{potential to interact} with opportunities as a
result of land-use mix and transport systems as originally defined by
{[}1{]}) have been widely applied within the transportation literature
and is increasingly discussed by planners {[}2--5{]}. An important
challenge in the identification of interventions that equitably
transform cities is the effective evaluation of \emph{trade-offs}:
cities are complex and dynamic ecologies, and advantaging one component
of the city can disadvantage another area, population, or sub-component
In this way, policy evaluation should take a \emph{systems} approach
{[}6{]}. One way of considering systems is from the perspective of the
\emph{finite}. As an illustration, consider the amount of transport
space within a city: the amount is typically finite so re-allocating
road space away from one mode directly impacts the performance of the
others (see the literature on road space reallocation e.g., {[}7{]}).
Evaluating policy impacts in the context of \emph{finity} provides a way
to contextualize the balance of trade-offs that the citizens of a city
should tolerate.

From the perspective of urban transport systems, location-based
accessibility measures have been used in the context of policy
evaluation. For instance, {[}8{]} assesses the transit accessibility
gains to healthcare and employment opportunities for disadvantaged
neighbourhood in Columbus, Ohio, USA after the transit system's
re-design and introduction of a rapid bus system. However, a limitation
of this study, like others that implement accessibility measures, is
they do not calculate results under a \emph{constrained} framework i.e.,
one of \emph{finity}. The citizens of Columbus should experience
quantitative accessibility gains - but is it at the expense of access
using other modes? As another example, {[}9{]} implements a modified
cumulative opportunity measure to assess differences between private
vehicle and transit system accessibility to jobs in Melbourne, but a
similar question remains: does the accessibility afforded to the private
vehicle using population come at the expense of accessibility losses to
transit users?

{[}8{]} and {[}9{]} both use \emph{non-competitive} accessibility
measures. There is a branch of location-based accessibility measures
that do incorporate the effect of competition for opportunities by the
population in the region. However, we argue that these existing methods
fall short in acknowledging the \emph{finity} of opportunities. For
instance, {[}10{]} applies a competitive measure, two-step floating
catchment approach (2SFCA), for the case of access to healthcare
services in Florida for both a multimodal network and a single modal
network. While the differences in modal access are discussed, the
question of how the advantage in access afforded by one mode over
another impacts access for different mode users is unanswered.

This question of how much one mode-using population can access at the
expense of another mode-using population is a pertinent equity question
in the evaluation of policy scenarios that are multimodal. For instance,
consider the impact of a low emission zone (LEZ). LEZ is a policy of
spatial and modal discrimination: the circulation of vehicles that are
excessively polluting are restricted in specific regions. In the
recognition that opportunities are finite, the implementation of a LEZ
explicitly reduces the access that the population using polluting
vehicles has to opportunities. This restriction allows the population
using other more sustainable modes to potentially have a higher level of
access than before the LEZ implementation. This evaluation is especially
urgent as LEZ are currently in effect in cities globally; their
reception has been mixed {[}11{]} and may be having negative impacts on
disadvantaged populations who have become mobility-restricted
{[}12,13{]}. Measures that evaluate the accessibility of modes given
both \emph{constrained} and \emph{competitive} considerations are
lacking in the literature, but are needed, to evaluate such policy
interventions impact on accessibility.

In {[}14{]}, we introduce spatial availability, a type of location-based
accessibility measure that is both \emph{constrained} and
\emph{competitive}. In this paper, we extend the spatial availability
measure into a multimodal framework and explore its use in answering the
question outlined: \emph{``given opportunities are finite, how many are
available to a given location depending on the mode used?''.} The answer
to this question quantifies how many opportunities can be accessed,
considering competition, for different modes. To foreground this
exploration, in Section 2, we discuss short falls of a few existing
location-based measures in comparison to spatial availability through a
synthetic example. In Section 3, the spatial availability of an
empirical example of the LEZ in the city of Madrid, Spain is calculated.
We demonstrate how the restriction of car circulation could have
impacted the spatial availability of opportunities for each
sub-population using transit, cycling and walking modes. In Section 4,
we provide concluding remarks on the strengths of the use of spatial
availability as a multimodal accessibility measure and potential future
uses in policy planning scenarios.

@@

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are a product of the relationship between land-use
and transport systems. Arguably the most commonly used location-based
measured are cumulative opportunity measures and weighted cumulative
opportunity measures {[}2{]}. These measures weight the opportunities
that can be potentially interacted with from origin \(i\) to destination
\(j\) based on some sort of travel cost function (e.g., travel time,
fare, travel distance) otherwise known as a travel impedance function
\(f^{m}(c^m_{ij})\). Many weighted cumulative opportunities (often
referred to as the gravity-based measure) originate from the measure
proposed by {[}1{]}, which can take the following multimodal form:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
have mode-specific travel costs \((c_{ij}^m)\) and travel impedance
functions \(f^m(\cdot)\).

The Hansen-type measure does not consider competition between modes nor
is it constrained. As an example, the work of {[}15{]} uses the
Hansen-type measure to measure the potential interaction with retail
locations using walking, public transit, and car modes \(m\). \(S_i^m\)
is the sum of retail locations \(j\) that can potentially be interacted
with under the travel impedance as calculated for each \(i\) and \(m\).
In other words, each \(i\) has three \(S_i\) values, one per \(m\). In
this work, they demonstrate that the car mode has the highest
\(S_i^{m=car}\) values in the majority of \(i\), i.e., populations using
a car can potentially interact with the most retail opportunities than
populations using other modes. However, the higher \(S_i^{m=car}\)
values are not a result of lower \(S_i^{m}\) values for other modes: it
is not assumed that car-using populations potentially accessing more
opportunities take away potential opportunities for other populations
within the measure (no consideration for competition). This measure is
also not constrained: there is no global maximum for \(S_i\) or
\(S_i^m\) values, they are presented as a population normalized
accessibility index. This makes the interpretation of the `potentially
interacted opportunities' relative to the region, making comparisons of
the results across different regions challenging.

However, opportunities in a region can be considered finite. There are
only so many school-seats, hospital capacity, jobs, etc., in a region
and if one person interacts with an opportunity at a given time, it is
taken. As such, if one person is advantaged and has the ability to reach
more opportunities through a lower travel-cost mode, than they have more
opportunities to potentially interact with more opportunities than other
people. From the other perspective, their are fewer opportunities left
to be potentially interacted with for populations using higher
travel-cost modes. In this way, populations using modes with a higher
travel impedance are at a higher access disadvantage than populations
using lower travel impedance modes. This recognition is the motivation
behind integrating \emph{competition} for opportunities within
multimodal accessibility measures. Arguably one of the most popular
competitive location-based accessibility measures is the two-step
floating catchment area (2SFCA) approach popularized by {[}16{]} who
simplified the approach proposed by {[}17{]} (with similar
considerations for competition in {[}18{]} and {[}19{]}).

The Shen-type accessibility measure's formulation is:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\). In this way, the
Shen-type measure can be understood as a ratio of the potential
opportunity supply over the potential demand for opportunities. The
measure considers competition, but it is \emph{non-constrained}. A score
of competitive potential accessibility associated is associated with
each location \(i\) for each mode \(m\), but there are no global
maximums. In other words, it is difficult to interpret the meaning of
differences in Shen-type accessibility scores between modes.

To illustrate, {[}20{]} calculates \(a_i^m\) to jobs for different
income-group populations in Shenzhen (China) using
\(m = \text{public transit}\) and \(m={car}\). They demonstrate that
\(i\) with low-income populations have lower \(a_i^m\) than \(i\) with
higher-income populations. Further, they demonstrate that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many \(i\), arguing that this may put \(i\) with lower-income
populations in a further disadvantage. \(a_i\) and/or \(a_i^m\) are used
to compare relative spatial differences in overall competitive
accessibility and modal competitive accessibility, but because there is
no global maximum, making it is difficult to interpret the significance
between differences in \(a_i^{m}\) values. Questions such as: what is
the impact that competition has on the difference in \(a_i^m\) values?
How does impact vary spatially? And what is the interpretation of this
difference? are left unanswered.

Spatial availability improves on previous multimodal accessibility
approaches as it considers \emph{competition} in the potential
interaction with opportunities in a \emph{constrained} framework (e.g.,
finite opportunities). This is done by considering: 1) competition
between mass effect (e.g., the advantage of sub-populations residing in
relatively low population-density and high opportunity-proximate areas)
and 2) competition between travel impedance (e.g., sub-populations with
relatively low travel-impedance) through a proportional allocation
mechanism. The following sub-section demonstrates how spatial
availability compares to the Hansen-type and Shen-type measures through
a synthetic example.

@@

In brief, we define the \emph{spatial availability} at \(i\) (\(V_{i}\))
as the proportion of all opportunities in the region \(O\) that are
allocated to location \(i\) from all opportunity destinations \(j\).
\(V_{i}\) is a value of how many opportunities are available to each
location \(i\) out of all the opportunities in the region. The general
formulation of spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}):

@@

The spatial availability measure is introduced in {[}14{]}. Spatial
availability's unique feature is the balancing factor \(F^t_{ij}\), a
proportional allocation mechanisms, that ensures the \(V_i\) calculated
for each \(i\) sums, across all \(i\) in the region, to equal the total
number of opportunities in the region. As such, spatial availability is
a \emph{competitive} and \emph{constrained} accessibility measure as
\(F^t_{ij}\) handles the number of opportunities in the region in a
finite way (proportional allocation). \(F^t_{ij}\) consists of two
components: a population-based balancing factor
\(F^p_{i} = \frac{P_i}{\sum_i P_i}\) and an impedance-based balancing
factor \(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that,
respectively, allocate opportunities to \(i\) in proportion to the size
of the population at \(i\) (the mass effect) and the cost of reaching
opportunities at \(j\) (the impedance effect).

\(F^p_{i}\) and \$ F\^{}c\_\{ij\}\$ are calculated for each \(i\) such
that they both equal 1 when summed across all \(i\) in the region (e.g.,
\(\sum_i F^p_{i} = 1\) and \(\sum_i F^c_{ij} = 1\)). These balancing
factors are combined multiplicatively to yield \(F^t_{ij}\) which
ensures that a proportion of the opportunities \(O_j\) are allocated to
each \(i\) accordingly. In other words, assuming a finite number of
opportunities in the region, \(F^t_{ij}\) proportionally allocates
\(O_j\) to each \(i\) such that the resulting \(V_i\) value represents
the number of opportunities \emph{spatially available} to the population
at \(i\). This value can be seen to represent spatial availability as it
is a proportion of the opportunities in the region (i.e.,
\(\sum_j O_j = \sum_i V_i\)).

The focus of this paper is to extend \(V_i\) for multimodal
applications. To do so, the balancing factors are reformulated to yield
a proportional value for the set of modes \(m\) used by populations at
each \(i\). As these factors are proportional, \(F^{pm}_{i}\) and
\(F^{cm}_{ij}\) can be summed up across each \(m\) at each \(i\) and
across all \(i\) to equal to 1. They are also similarly combined
multiplicatively to obtain their joint effect, represented as the
combined balancing factor \(F^{tm}_{ij}\) detailed in Equation
(\ref{eq:multimodal-balancing-factors}).

@@

\begin{itemize}
\tightlist
\item
  The population balancing factor for each \(m\) at each \(i\) is
  \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\)
\item
  The cost of travel balancing factor for each \(m\) at \(i\) is
  \(F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m} \sum_{i} f(c_{ij}^m)}\)
\end{itemize}

@@

\hypertarget{synthetic-example}{%
\subsection{Synthetic example}\label{synthetic-example}}

Consider the following: Figure \ref{fig:Fig1} depicts a region with
population and jobs at three population centers (\(A\), \(B\), \(C\))
and three employment centers (\(1\), \(2\), \(3\)). The population at
each population center is divided into two sub-populations, one using a
faster mode \(z\) and another using a slower mode \(x\), to travel to
employment centers. Population center \(A\) is Suburban: it is closest
to its own relatively large employment center at \(1\), close to the
Urban's equally large employment center \(2\), and has a population that
is smaller than the Urban \(B\) and larger than the Satellite \(C\).
\(B\) has the largest \(x\)-using population, followed by then \(A\),
then \(C\). This synthetic example was inspired by the single-mode
example used in {[}17{]} and reconfigured in {[}14{]}.

@@

From the perspective of access to a \emph{finite} amount of
opportunities in the region (\(210,000\) jobs), the sub-population that
is most proximate to jobs, furthest from densely populated centers, and
is using the lowest travel-cost mode \(z\) can potentially access the
most job opportunities. This appears to be the sub-population at \(A\)
using \(z\). From the other perspective, sub-populations located in
opposite conditions (i.e., further away from jobs, close to dense
populations, and using \(x\)) are at a relative job opportunity access
\emph{disadvantage}. From the perspective of inequities, the competition
for opportunities between different mode-using populations matters as it
reflects how well the land-use and transport system serves (or doesn't
serve) them.

@@

The calculated \(S_i^m\), \(a_i^m\) and \(V_i^m\) accessibility values
for each \(i\) and \(m\) are shown in the middle three columns and are
aggregated for each \(i\) in the final two columns in Table 1 . We use a
negative exponential impedance function
\(f(c_{ij}) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
\(x\) and \(z\) modes for all accessibility measures calculations.

The Hansen-type measure \(S_i^m\) is presented for each origin and mode
in third column of Table 1 . For all \(i\), the \(z\)-using
sub-population has higher \(S_i^m\) values than the \(x\)-using
sub-populations. Additionally, \(S_i^m\) is equal for both mode-using
populations in \(A\) and \(B\). This is the case because \(S_i^m\) does
not consider \emph{competition}, it only relies on reflecting the count
of opportunities that may be interacted with as a product of
\(f^m(c_{ij}^m)\). Recall, populations in \(A\) and \(B\) have the same
travel impedance to employment centers \(1\), \(2\) and \(3\) (either
15, 30, or 100 minutes using \(x\) or 10, 25, or 80 minutes using
\(z\)). As such, these the calculated \(S_i^m\) values are the same for
both \(A\) and \(B\). Furthermore, the total sum of \(S_i^m\) in the
region is equal to 150,570.2. This value is difficult to interpret: it
represents the weighted sum of opportunities that may be interacted with
within the region based on travel impedance. It cannot be interpreted as
any sort of benchmark since the measure is \emph{non-constrained}. To
connect this example to literature, \(S_i^m\) is calculated in the work
of {[}15{]}; they compare differences in \(S_i^m\) values between modes
in a relative and comparative sense, but make no further interpretation
of the \(S_i^m\) values.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multimodal spatial availability: a singly-constrained
measure of accessibility considering multiple
modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anastasia Soukhov\textsuperscript{\getAff{McMaster}}\textsuperscript{*},
Javier Tarriño-Ortiz\textsuperscript{\getAff{Madrid}},
Julio A. Soria-Lara\textsuperscript{\getAff{Madrid}},
Antonio Páez\textsuperscript{\getAff{McMaster}}\\
\bigskip
\textbf{\getAff{McMaster}}Department of Earth, Environment and Society,
McMaster University, Hamilton, Canada\\
\textbf{\getAff{Madrid}}Centro de Investigación del Transporte
(TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain\\
\bigskip
* Corresponding author: soukhoa@mcmaster.ca (AS)\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Recent research has tried to address the way opportunities are counted
in accessibility analysis. In conventional accessibility measures,
opportunities are often multiply counted, which leads to values of
accessibility that are difficult to interpret. Constraining the
calculations to match a known quantity ensures that the measurements sum
up to a predetermined quantity (i.e., the total number of
opportunities), and so each value can be meaningfully related to this
total. A recent such effort is spatial availability, a
singly-constrained accessibility measure. In this paper we extend
spatial availability for use in the case of multiple modes or, more
generally, heterogeneous population segments with distinct travel
behaviors. After deriving a multimodal version of spatial availability
We proceed to illustrate its features using a synthetic example. Next,
we apply it to an empirical example in Madrid. We conclude the paper
with suggestions for future research.

@@

Accessibility is a key concept in the analysis of land use and
transportation systems {[}e.g., 1,2,3{]}, and one that is coming of age
from the perspective of planning too {[}see, \emph{inter alia},
4,5--8{]}. Beginning with the work of {[}1{]}, accessibility measures
have been widely used to evaluate the efficiency of transportation
systems when combined with the distribution of opportunities in space
{[}e.g., {]}. As such, it is a holistic measure of spatial systems that
measures the ease of reaching destinations {[}9,10{]}.

In practice, the most common form of accessibility measure is based on
the gravity model. These measures are sums of weighted opportunities
around a focal point (i.e., a potential origin), based on how expensive
it is to reach them. Recent research in accessibility analysis has paid
attention to the way opportunities are counted in the pertinent
calculations. Conventionally, the sums are not constrained, which means
that the same opportunity can enter the sum for different origins.
Counting the same opportunity multiple times treats it as if it was
inexhaustible. But opportunities in general are not inexhaustible, and
in fact some of them are by definition exclusive: for example, once a
job is taken up by someone in the population, the same job is no longer
available for any other person to take. Other types of opportunities are
at the very least subject to congestion: multiple people can avail
themselves of the services of the same family doctor, but the more
people who do, the more congested the service will be.

The issue of congestion was the motivation for the development of
floating catchment area approaches in accessibility analysis
{[}11,12{]}. While these approaches purport to account for congestions,
{[}13{]} demonstrate that in general they do not solve the issue of
multiple counting of opportunities, thus leading to biases in the
calculation of total demand and supply, sometimes inflating them, other
times deflating them. In response to this, recent research has paid
closer attention to the way opportunities are counted in accessibility
analysis. {[}13{]}, for example, tackle floating catchment area methods
and introduce a normalization of the impedance matrix to allocate the
population and then the level of service proportionally. More recently,
{[}14{]} introduced a singly-constrained measure of accessibility,
called spatial availability, that employs a similar, but more
sophisticated proportional allocation mechanism. The work of these
authors shows that floating catchment area methods can be seen as
singly-constrained accessibility measures, and improve on them by
guaranteeing that each opportunity is counted only once - in other
words, treating opportunties as \emph{finite}. The proportional
allocation of spatial availability constrains the calculations to match
a known quantity, therefore ensuring that the measurements sum up to a
predetermined quantity (i.e., the total number of opportunities), and so
each value can be meaningfully related to this total.

A limitation of spatial availability as introduced by {[}14{]} is that
it was developed for the case of a homogeneous population, for example
for the case of a single mode of transportation. However, the finity of
the opportunities makes the analysis of heterogeneous populations very
relevant. In the case of multiple modes of transportation, people who
travel by slow modes (e.g., active modes) can usually reach fewer
opportunities than people who travel by faster modes and whose range is
typically far wider (e.g., car). This implies that slower travelers will
often face increased competition for local opportunities from travelers
who can reach said opportunities from farther afield.

The objective of this paper is to address this limitation of spatial
availability. Our primary motivation is to extend spatial availability
for the case of multimodal accessibility, but this is in fact just one
example of heterogeneous populations (i.e., travel by different modes).
The method itself can easily accommodate other forms of heterogeneity,
for example variations in travel behavior between older and younger
adults {[}e.g., {[}15{]}; {]}, the propensity of older adults to use
different modes of transportation {[}16{]}, the usually shorter trip
lengths of children compared to grown-ups {[}e.g., 17{]}, or the more
limited travel ranges of single parents {[}e.g., 18{]}.

The paper rest of this paper is organized as follows. In Section we use
a synthetic examples with multiple modes to illustrate some relevant
issues; this helps us motivate the derivation of the new spatial
availability expressions. This is followed in Section 3 by an empirical
example using data from the city of Madrid after the implementation of
its Low Emission Zones. Data for this example comes from the city's 2018
travel survey. The example shows the differences in spatial availability
within and outside the LEZ for travellers using different modes, namely
car, transit, cycling and walking. In Section 4, we provide concluding
remarks on the strengths of the use of spatial availability as a
multimodal accessibility measure, and discuss potential future uses in
policy planning scenarios as well as directions for future research.

@@

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are summary measures of the relationship between
land-use and transport systems. Arguably, the most commonly used are
measures based on the gravity model {[}19{]}, of which cumulative
opportunity measures and weighted cumulative opportunity measures are
particular forms {[}5{]}. These measures assign a weight to
opportunities based on how easy it is to reach them. Given an origin
(\(i\)) and a destination (\(j\)), an impedance function
\(f^{m}(c^m_{ij})\) converts the cost of travel (e.g., time, money,
generalized cost) into a score that represents the propensity for
interaction. These measures originate from that proposed by {[}1{]},
which can take the following form in the multimodal case:
\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
have mode-specific travel costs (\((c_{ij}^m)\)) and/or travel impedance
functions (\(f^m(\cdot)\)).

The Hansen-type measure is not constrained, and as a result does not
consider the opportunities as finite. As an example, the work of
{[}20{]} uses the Hansen-type measure to measure the potential
interaction with retail locations using walking, public transit, and car
modes as \(m\). \(S_i^m\) is the sum of retail locations \(j\) that can
potentially be reached under the travel impedance as calculated for each
\(i\) and \(m\). In other words, each \(i\) has three \(S_i^m\) values,
one per \(m\). In this work, they demonstrate that the car mode has the
highest \(S_i^{m}\) values in the majority of \(i\), i.e., populations
using a car can potentially reach more retail opportunities than
populations using other modes. However, higher \(S_i^{m}\) values for
car do not affect the values of \(S_i^{m}\) for other modes: in effect,
each modes is analyzed as if the others did not exist, which would be
reasonable for an inexhaustible opportunity, but is more questionable
when the opportunity is subject to congestion. Since the measure is
measure is also not constrained, and each opportunity can be and often
is counted multiple times within and between modes, the sum of
accessibility is not a meaningful quantity. To improve interpretability,
\(S_i^m\) values are presented as standardized accessibility index in
this work, from the lowest value of 0 to the highest value of 1. This
presentation or discussion of Hansen-type accessibility measures is
common in the literature {[}e.g., in 21,and 22{]}.

However, this representation of results can make the interpretation of
results challenging \emph{because} values are region-relative and
lacking numeric meaning. Zones will always have values from between 0 to
1 represented, but is a zone with a low value for pedestrian-modes and a
high value for car-modes notable? And if notable, what does the
difference in these standardized values tangibly mean for planners? By
how much should transport systems and land-use configurations be changed
to improve conditions? And in what way can these existing methods be
used to track progress over time? Or between regions? This is
challenging to track, since certain values will always be relatively
`low' and `high'. Alternatively, modal accessibility calculated using
the Hansen-type measure is also presented without standardization, as in
the work by {[}23{]}. Often, the meaning of the accessibility value
itself is not interpreted, as it is difficult to state what having
access to 30,000 employment opportunities by car but having 10,000
employment opportunity by transit tangibly signifies.

However, once opportunities in a region are interpreted as finite, a new
meaning is taken on. As considered in the long tradition of
accessibility work, capacity of opportunities is limited and thus is
subject to competition by population {[}11,14,24--28{]}. There are only
so many school-seats, hospital capacity, employment opportunities, etc.,
in a region and if one person interacts with an opportunity at a given
time, it is taken: the supply of an opportunity and the demand for that
opportunity are two components of accessibility. The competition
consideration is especially important in the case of multiple modes. For
instance, people in a zone who are advantaged with a relatively low
travel-cost mode have the ability to potentially interact with more
opportunities than other people. Due to this advantage, through the
perspective of finite opportunities, there are fewer opportunities left
to be potentially interacted for everyone else, especially zones with
populations using higher travel-cost modes. This recognition is the
motivation behind integrating \emph{competition} for opportunities
within modal accessibility measures.

Arguably, one of the most popular competitive location-based
accessibility measures is the two-step floating catchment area (2SFCA)
approach popularized by {[}12{]} who simplified the approach proposed by
{[}11{]} (with similar considerations for competition in {[}28{]} and
{[}29{]}). The Shen-type accessibility measure's formulation is:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
variables are repeated in the Hansen-type measure. The Shen-type measure
per mode (\(a_i^m\)) can be understood as a ratio of the travel
impedance-weighted supply of opportunities for the \(m\)-using
population in \(i\) over the ravel impedance-weighted demand for
opportunities. In this way, it considers competition but the measure
remains \emph{non-constrained} meaning resulting values are not
associated with global maximums. In other words, it is difficult to
interpret the meaning of differences in Shen-type accessibility scores
between modes in similar ways as reflected for Hansen-type measures.

To illustrate, {[}30{]} calculates \(a_i^m\) to jobs for different
income-group populations in Shenzhen, China using
\(m = \text{public transit}\) and \(m={car}\). They demonstrate that
\(i\) with low-income populations have lower \(a_i^m\) than \(i\) with
higher-income populations. Further, they demonstrate that
\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
at many \(i\), arguing that this may put \(i\) with lower-income
populations in a further disadvantage. \(a_i\) and/or \(a_i^m\) are used
to compare relative spatial differences in overall competitive
accessibility and multimodal competitive accessibility, but because
there is no global maximum, making it is difficult to interpret the
significance between differences in \(a_i^{m}\) values. Questions such
as: what is the impact that competition has on the difference in
\(a_i^m\) values? How does impact vary spatially? And what is the
interpretation of this difference? are left unaddressed.

Spatial availability improves on previously discussed multimodal
accessibility approaches using the Hansen-type measure and the Shen-type
measure by taking into account \emph{competition} in the potential
interaction with opportunities in a \emph{constrained} framework (e.g.,
finite opportunities). This is done by considering: 1) competition
between mass effect (e.g., the advantage of sub-populations residing in
relatively low population-density and high opportunity-proximate areas)
and 2) competition between travel impedance (e.g., sub-populations with
relatively low travel-impedance) through a proportional allocation
mechanism. The following sub-section demonstrates how spatial
availability compares to the Hansen-type and Shen-type measures through
a synthetic example.

@@

In brief, we define the spatial availability at \(i\) (\(V_{i}\)) as the
proportion of all opportunities in the region \(O\) that are allocated
to location \(i\) from all opportunity destinations \(j\). \(V_{i}\) is
a value of how many opportunities are available to each location \(i\)
out of all the opportunities in the region. The general formulation of
spatial availability \(V_{i}\) is shown in Equation
(\ref{eq:spatial-availability-general}):

@@

The spatial availability measure is introduced in {[}14{]}. Spatial
availability's unique feature is the balancing factor \(F^t_{ij}\), a
proportional allocation mechanisms, that ensures the \(V_i\) calculated
for each \(i\) sums, across all \(i\) in the region, to equal the total
number of opportunities in the region. As such, spatial availability is
a \emph{competitive} and \emph{constrained} accessibility measure as
\(F^t_{ij}\) handles the number of opportunities in the region in a
finite way (proportional allocation). \(F^t_{ij}\) consists of two
components: a population-based balancing factor
\(F^p_{i} = \frac{P_i}{\sum_i P_i}\) and an impedance-based balancing
factor \(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that,
respectively, allocates opportunities to \(i\) in proportion to the size
of the population at \(i\) (the mass effect) and the cost of reaching
opportunities at \(j\) (the impedance effect).

\(F^p_{i}\) and \$ F\^{}c\_\{ij\}\$ are calculated for each \(i\) such
that they both equal 1 when summed across all \(i\) in the region (e.g.,
\(\sum_i F^p_{i} = 1\) and \(\sum_i F^c_{ij} = 1\)). These balancing
factors are combined multiplicatively to yield \(F^t_{ij}\) which
ensures that a proportion of the opportunities \(O_j\) are allocated to
each \(i\) accordingly. In other words, assuming a finite number of
opportunities in the region, \(F^t_{ij}\) proportionally allocates
\(O_j\) to each \(i\) such that the resulting \(V_i\) value represents
the number of opportunities \emph{spatially available} to the population
at \(i\). Each zonal value is a proportion of the opportunities in the
region (i.e., \(\sum_j O_j = \sum_i V_i\)).

The focus of this paper is to extend \(V_i\) for the measurement of
multimodal accessibility applications. To do so, the balancing factors
are reformulated to yield a proportional value for the set of modes
\(m\) used by populations at each \(i\). As these factors are
proportional, \(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be summed across
each \(m\) at each \(i\) and then across all \(i\) to equal to 1. They
are also similarly combined multiplicatively to obtain their joint
effect, represented as the combined balancing factor \(F^{tm}_{ij}\)
detailed in Equation (\ref{eq:multimodal-balancing-factors}).

@@

\begin{itemize}
\tightlist
\item
  The population balancing factor for each \(m\) at each \(i\) is
  \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\)
\item
  The cost of travel balancing factor for each \(m\) at \(i\) is
  \(F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}\)
\end{itemize}

@@

\hypertarget{an-illustrative-synthetic-example}{%
\subsection{An illustrative synthetic
example}\label{an-illustrative-synthetic-example}}

Consider the following Figure \ref{fig:Fig1}, it depicts a region with
population and jobs at three population centers (\(A\), \(B\), \(C\))
and three employment centers (\(1\), \(2\), \(3\)). The population at
each population center is divided into two sub-populations, one using a
faster mode \(z\) and another using a slower mode \(x\), to travel to
employment centers. Population center \(A\) is Suburban: it is closest
to its own relatively large employment center at \(1\), close to the
Urban's equally large employment center \(2\), and has a population that
is smaller than the Urban \(B\) and larger than the Satellite \(C\).
\(B\) has the largest \(x\)-using population, followed by then \(A\),
then \(C\). This synthetic example was inspired by the single-mode
example used in {[}11{]} and reconfigured in {[}14{]}.

@@

From the perspective of access to a \emph{finite} amount of
opportunities in the region (\(210,000\) jobs), the sub-population that
is most proximate to jobs, furthest from densely populated centers, and
is using the lowest travel-cost mode \(z\) can potentially access the
most job opportunities. This appears to be the sub-population at \(A\)
using mode \(z\). Sub-populations located in opposite conditions (i.e.,
further away from jobs, close to dense populations, and using \(x\)) are
at a relative job opportunity access \emph{disadvantage}. The
competition for opportunities between different mode-using populations
matters from the perspective of inequities as it reflects how well the
land-use and transport system serves (or doesn't serve) certain
populations.

@@

The calculated \(S_i^m\), \(a_i^m\) and \(V_i^m\) accessibility values
for each \(i\) and \(m\) are shown in the middle three columns and are
aggregated for each \(i\) in the final two columns in Table 1 . We use a
negative exponential impedance function
\(f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
\(x\) and \(z\) modes for all accessibility measures calculations.

The Hansen-type measure \(S_i^m\) is presented for each origin and mode
in third column of Table 1 . For all \(i\), the \(z\)-using
sub-population has higher \(S_i^m\) values than the \(x\)-using
sub-populations. Additionally, \(S_i^m\) is equal for both mode-using
populations in \(A\) and \(B\). This is the case since \(S_i^m\) does
not consider \emph{competition}, it only relies on reflecting the count
of opportunities that may be interacted with as a product of
\(f^m(c_{ij}^m)\). Recall, populations in \(A\) and \(B\) have the same
travel impedance to employment centers \(1\), \(2\) and \(3\) (either
15, 30, or 100 minutes using \(x\) or 10, 25, or 80 minutes using
\(z\)). As such, these the calculated \(S_i^m\) values are the same for
both \(A\) and \(B\). Furthermore, the total sum of \(S_i^m\) in the
region is equal to 150,570.2. This value is difficult to interpret: it
represents the weighted sum of opportunities that may be interacted with
within the region based on travel impedance. It cannot be interpreted as
any sort of benchmark since the measure is \emph{non-constrained}. To
connect this example to literature, \(S_i^m\) is calculated in the work
of {[}20{]}; they compare differences in \(S_i^m\) values between modes
in a relative and comparative sense, but make no further interpretation
of the \(S_i^m\) values.
